---
title: "Section 2 Continuous Probability"
author: "kevinoliva26"
date: "5/14/2021"
output: html_document
---
##Continuous Probability

The CDF is the probability distribution function for continuous variables. For example, to determine the probability that a male student is taller than 70.5 inches given a vector of male heights  x , we can use the CDF:

*Pr(x>70.5)=1−Pr(x≤70.5)=1−F(70.5) *

The probability that an observation is in between two values  a,b  is  F(b)−F(a) .
```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)

F <- function(a) mean(x <= a)
1 - F(70)    # probability of male taller than 70 inches
```

##Theoretical Distribution

<<<<<<< HEAD
=======
*pnorm(a, avg, s)* gives the value of the cumulative distribution function F(a) for the normal distribution defined by average avg and standard deviation s.

We say that a random quantity is normally distributed with average avg and standard deviation s if the approximation pnorm(a, avg, s) holds for all values of a.

If we are willing to use the normal approximation for height, we can estimate the distribution simply from the mean and standard deviation of our values.
If we treat the height data as discrete rather than categorical, we see that the data are not very useful because integer values are more common than expected due to rounding. This is called discretization.
With rounded data, the normal approximation is particularly useful when computing probabilities of intervals of length 1 that include exactly one integer.

Code: Using pnorm() to calculate probabilities
```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)
head(x)

#We can estimate the probability that a male is taller than 70.5 inches using:

1 - pnorm(70.5, mean(x), sd(x))
```

Code: Discretization and the normal approximation
```{r}
# plot distribution of exact heights in data
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer
mean(x <= 68.5) - mean(x <= 67.5)
mean(x <= 69.5) - mean(x <= 68.5)
mean(x <= 70.5) - mean(x <= 69.5)

# probabilities in normal approximation match well
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))
pnorm(70.5, mean(x), sd(x)) - pnorm(69.5, mean(x), sd(x))

# probabilities in actual data over other ranges don't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))
```
>>>>>>> f0bd94c... Added content


##Probability Density

In R, the probability density function for the normal distribution is given by *dnorm()*. We will see uses of dnorm() in the future.
<<<<<<< HEAD
Note that *dnorm()* gives the density function and *pnorm()* gives the distribution function, which is the integral of the density function.
=======
Note that *dnorm()* gives the density function and *pnorm()* gives the distribution function, which is the integral of the density function.

##Plotting the probability density for the normal distribution

We can use dnorm() to plot the density curve for the normal distribution. dnorm(z) gives the probability density f(z) of a certain z-score, so we can draw a curve by calculating the density over a range of possible values of z.

First, we generate a series of z-scores covering the typical range of the normal distribution. Since we know 99.7% of observations will be within −3≤z≤3, we can use a value of z slightly larger than 3 and this will cover most likely values of the normal distribution. Then, we calculate f(z), which is dnorm() of the series of z-scores. Last, we plot z against f(z).

```{r}
library(tidyverse)
x <- seq(-4, 4, length = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x, f)) +
    geom_line()
```

Note that dnorm() gives densities for the standard normal distribution by default. Probabilities for alternative normal distributions with mean mu and standard deviation sigma can be evaluated with:

*dnorm(z, mu, sigma)*

##Monte Carlo Simulations

rnorm(n, avg, s) generates n random numbers from the normal distribution with average avg and standard deviation s.
By generating random numbers from the normal distribution, we can simulate height data with similar properties to our dataset. Here we generate simulated height data using the normal distribution.

Code: Generating normally distributed random numbers
```{r}
# define x as male heights from dslabs data
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)

# generate simulated height data using normal distribution - both datasets should have n observations
n <- length(x)
avg <- mean(x)
s <- sd(x)
simulated_heights <- rnorm(n, avg, s)

# plot distribution of simulated_heights
data.frame(simulated_heights = simulated_heights) %>%
    ggplot(aes(simulated_heights)) +
    geom_histogram(color="black", binwidth = 2)
```

Code: Monte Carlo simulation of tallest person over 7 feet
```{r}
B <- 10000
tallest <- replicate(B, {
    simulated_data <- rnorm(800, avg, s)    # generate 800 normally distributed random heights
    max(simulated_data)    # determine the tallest height
})
mean(tallest >= 7*12)    # proportion of times that tallest person exceeded 7 feet (84 inches)
```

##Other Continuous Distributions

You may encounter other continuous distributions (Student t, chi-squared, exponential, gamma, beta, etc.).
R provides functions for *density (d)*, *quantile (q)*, *probability distribution (p) * and r*andom number generation (r) * for many of these distributions.
Each distribution has a matching abbreviation (for example, norm() or t()) that is paired with the related function abbreviations *(d, p, q, r)* to create appropriate functions.
For example, use rt() to generate random numbers for a Monte Carlo simulation using the Student t distribution.

Code: Plotting the normal distribution with dnorm
```{r}
x <- seq(-4, 4, length.out = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x,f)) +
    geom_line()
```

##Questions 1 and 2: ACT scores, part 1


```{r}
set.seed(16, sample.kind = "Rounding")
act_scores <- rnorm(10000, 20.9,5.7)

#What is the mean of act_scores
mean(act_scores)

#What is the standard deviation of act_scores?
sd(act_scores)

#A perfect score is 36 or greater (the maximum reported score is 36). In act_scores, how many perfect scores are there out of 10,000 simulated tests?
sum(act_scores >=36)

#In act_scores, what is the probability of an ACT score greater than 30?
mean(act_scores > 30)

#In act_scores, what is the probability of an ACT score less than or equal to 10?
mean(act_scores <= 10)

```

