---
title: "Section 1 Discrete Probability"
author: "kevinoliva26"
date: "5/14/2021"
output: html_document
---
##1.1 Introduction to Discrete Probability

Monte Carlo simulations model the probability of different outcomes by repeating a random process a large enough number of times that the results are similar to what would be observed if the process were repeated forever.
The *sample()* function draws random outcomes from a set of options.
The *replicate()* function repeats lines of code a set number of times. It is used with sample() and similar functions to run Monte Carlo simulations.

```{r}
beads <- rep(c("red", "blue"), times = c(2,3))    # create an urn with 2 red, 3 blue
beads    # view beads object
sample(beads, 1)    # sample 1 bead at random
B <- 10000    # number of times to draw 1 bead
events <- replicate(B, sample(beads, 1))    # draw 1 bead, B times
tab <- table(events)    # make a table of outcome counts
tab    # view count table
prop.table(tab)    # view table of outcome proportions
#Other whay to do it is with the replace argument
events <- sample(beads, B, replace = TRUE)
prop.table(table(events))
```

to ensure that results are exactly the same every time you run them, you can set R’s random number generation seed to a specific number with *set.seed()*

Conditional probabilities compute the probability that an event occurs given information about dependent events. For example, the probability of drawing a second king given that the first draw is a king is:
Pr(Card 2 is a king∣Card 1 is a king)=3/51 
If two events  A  and  B  are independent,  Pr(A∣B)=Pr(A) .
To determine the probability of multiple events occurring, we use the multiplication rule.

The multiplication rule for independent events is:

*Pr(A and B and C)=Pr(A)×Pr(B)×Pr(C) *
The multiplication rule for dependent events considers the conditional probability of both events occurring:

*Pr(A and B)=Pr(A)×Pr(B∣A) *
We can expand the multiplication rule for dependent events to more than 2 events:

*Pr(A and B and C)=Pr(A)×Pr(B∣A)×Pr(C∣A and B)*

##1.2 Combinations and Permutations

*paste()* joins two strings and inserts a space in between.
expand.grid() gives the combinations of 2 vectors or lists.
*permutations(n,r)* from the gtools package lists the different ways that r items can be selected from a set of n options when order matters.
*combinations(n,r)* from the gtools package lists the different ways that r items can be selected from a set of n options when order does not matter.

Introducing paste() and expand.grid()
```{r}
# joining strings with paste
number <- "Three"
suit <- "Hearts"
paste(number, suit)
# joining vectors element-wise with paste
paste(letters[1:5], as.character(1:5))
# generating combinations of 2 vectors with expand.grid
expand.grid(pants = c("blue", "black"), shirt = c("white", "grey", "plaid"))
```

Generating a deck of cards
```{r}
suits <- c("Diamonds", "Clubs", "Hearts", "Spades")
numbers <- c("Ace", "Deuce", "Three", "Four", "Five", "Six", "Seven", "Eight", "Nine", "Ten", "Jack", "Queen", "King")
deck <- expand.grid(number = numbers, suit = suits)
deck <- paste(deck$number, deck$suit)
# probability of drawing a king
kings <- paste("King", suits)
mean(deck %in% kings)
```

Permutations and combinations
```{r}
library(gtools)
permutations(5,2)    # ways to choose 2 numbers in order from 1:5
all_phone_numbers <- permutations(10, 7, v = 0:9)
n <- nrow(all_phone_numbers)
index <- sample(n, 5)
all_phone_numbers[index,]
permutations(3,2)    # order matters
combinations(3,2)    # order does not matter
```

Probability of drawing a second king given that one king is drawn
```{r}
hands <- permutations(52,2, v = deck)
first_card <- hands[,1]
second_card <- hands[,2]
sum(first_card %in% kings)
sum(first_card %in% kings & second_card %in% kings) / sum(first_card %in% kings)
```

Probability of a natural 21 in blackjack
```{r}
aces <- paste("Ace", suits)
facecard <- c("King", "Queen", "Jack", "Ten")
facecard <- expand.grid(number = facecard, suit = suits)
facecard <- paste(facecard$number, facecard$suit)
hands <- combinations(52, 2, v=deck) # all possible hands
# probability of a natural 21 given that the ace is listed first in `combinations`
mean(hands[,1] %in% aces & hands[,2] %in% facecard)
# probability of a natural 21 checking for both ace first and ace second
mean((hands[,1] %in% aces & hands[,2] %in% facecard)|(hands[,2] %in% aces & hands[,1] %in% facecard))
```

Monte Carlo simulation of natural 21 in blackjack
```{r}
# code for one hand of blackjack
hand <- sample(deck, 2)
hand
# code for B=10,000 hands of blackjack
B <- 10000
results <- replicate(B, {
    hand <- sample(deck, 2)
    (hand[1] %in% aces & hand[2] %in% facecard) | (hand[2] %in% aces & hand[1] %in% facecard)
})
mean(results)
```

##The Birthday Problem

*duplicated()* takes a vector and returns a vector of the same length with TRUE for any elements that have appeared previously in that vector.
We can compute the probability of shared birthdays in a group of people by modeling birthdays as random draws from the numbers 1 through 365. We can then use this sampling model of birthdays to run a Monte Carlo simulation to estimate the probability of shared birthdays.

```{r}
# checking for duplicated bdays in one 50 person group
n <- 50
bdays <- sample(1:365, n, replace = TRUE)    # generate n random birthdays
any(duplicated(bdays))    # check if any birthdays are duplicated
# Monte Carlo simulation with B=10000 replicates
B <- 10000
results <- replicate(B, {    # returns vector of B logical values
    bdays <- sample(1:365, n, replace = TRUE)
    any(duplicated(bdays))
})
mean(results)    # calculates proportion of groups with duplicated bdays
```

##Sapply

The function *sapply(x, f)* allows any other function f to be applied element-wise to the vector x

Note that the function body of compute_prob() is the code that we wrote in the previous video. If we write this code as a function, we can use sapply() to apply this function to several values of n.

Function for birthday problem Monte Carlo simulations
```{r}
# function to calculate probability of shared bdays across n people
compute_prob <- function(n, B = 10000) {
	same_day <- replicate(B, {
    	bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}
n <- seq(1, 60)
```

Code: Element-wise operation over vectors and sapply
```{r}
x <- 1:10
sqrt(x)    # sqrt operates on each element of the vector
y <- 1:10
x*y    # * operates element-wise on both vectors
compute_prob(n)    # does not iterate over the vector n without sapply
x <- 1:10
sapply(x, sqrt)    # this is equivalent to sqrt(x)
prob <- sapply(n, compute_prob)    # element-wise application of compute_prob to n
plot(n, prob)
```

Code: Computing birthday problem probabilities with sapply
```{r}
# function for computing exact probability of shared birthdays for any n
exact_prob <- function(n){
    prob_unique <- seq(365, 365-n+1)/365   # vector of fractions for mult. rule
    1 - prod(prob_unique)    # calculate prob of no shared birthdays and subtract from 1
}
# applying function element-wise to vector of n values
eprob <- sapply(n, exact_prob)
# plotting Monte Carlo results and exact probabilities on same graph
plot(n, prob)    # plot Monte Carlo results
lines(n, eprob, col = "red")    # add line for exact prob
```

##How Many Monte Carlo Experiments are Enough?

The larger the number of Monte Carlo replicates  B , the more accurate the estimate.
Determining the appropriate size for  B  can require advanced statistics.
One practical approach is to try many sizes for  B  and look for sizes that provide stable estimates.

This code runs Monte Carlo simulations to estimate the probability of shared birthdays using several B values and plots the results. When B is large enough that the estimated probability stays stable, then we have selected a useful value of B
```{r}
B <- 10^seq(1, 5, len = 100)    # defines vector of many B values
compute_prob <- function(B, n = 22){    # function to run Monte Carlo simulation with each B
	same_day <- replicate(B, {
    	bdays <- sample(1:365, n, replace = TRUE)
        any(duplicated(bdays))
    })
    mean(same_day)
}
prob <- sapply(B, compute_prob)    # apply compute_prob to many values of B
plot(log10(B), prob, type = "l")    # plot a line graph of estimates 
```

Celtics Excerise
```{r}
# This line of sample code simulates four random games where the Celtics either lose or win. Each game is independent of other games.
simulated_games <- sample(c("lose","win"), 4, replace = TRUE, prob = c(0.6, 0.4))
# The variable 'B' specifies the number of times I want the simulation to run. I will run the Monte Carlo simulation 10,000 times.
B <- 10000
# Use the `set.seed` function to make sure my answer matches the expected result after random sampling.
set.seed(1)
# Create an object called `celtic_wins` that first replicates the sample code generating the variable called `simulated_games` for `B` iterations and then tallies the number of simulated series that contain at least one win for the Celtics.
celtic_wins <- replicate(B, {simulated_games <- sample(c("lose","win"), 4, replace = TRUE, prob = c(0.6, 0.4)); any(simulated_games == c("win"))})  
# Calculate the frequency out of B iterations that the Celtics won at least one game.  
mean(celtic_wins) 
#> [1] 0.8757
```

##The Addition Rule

The addition rule states that the probability of event  A  or event  B  happening is the probability of event  A  plus the probability of event  B  minus the probability of both events  A  and  B  happening together.
Pr(A or B)=Pr(A)+Pr(B)−Pr(A and B)

##Monty Hall Problem

In the Monty Hall problem, contestants choose one of three doors that may contain a prize. Then, one of the doors that was not chosen by the contestant and does not contain a prize is revealed. The contestant can then choose whether to stick with the original choice or switch to the remaining unopened door.
Although it may seem intuitively like the contestant has a 1 in 2 chance of winning regardless of whether they stick or switch, Monte Carlo simulations demonstrate that the actual probability of winning is 1 in 3 with the stick strategy and 2 in 3 with the switch strategy.

Code: Monte Carlo simulation of stick strategy
```{r}
B <- 10000
stick <- replicate(B, {
	doors <- as.character(1:3)
	prize <- sample(c("car","goat","goat"))    # puts prizes in random order
	prize_door <- doors[prize == "car"]    # note which door has prize
	my_pick  <- sample(doors, 1)    # note which door is chosen
	show <- sample(doors[!doors %in% c(my_pick, prize_door)],1)    # open door with no prize that isn't chosen
	stick <- my_pick    # stick with original door
	stick == prize_door    # test whether the original door has the prize
})
mean(stick)    # probability of choosing prize door when sticking
```

Code: Monte Carlo simulation of switch strategy
```{r}
switch <- replicate(B, {
	doors <- as.character(1:3)
	prize <- sample(c("car","goat","goat"))    # puts prizes in random order
	prize_door <- doors[prize == "car"]    # note which door has prize
	my_pick  <- sample(doors, 1)    # note which door is chosen first
	show <- sample(doors[!doors %in% c(my_pick, prize_door)], 1)    # open door with no prize that isn't chosen
	switch <- doors[!doors%in%c(my_pick, show)]    # switch to the door that wasn't chosen first or opened
	switch == prize_door    # test whether the switched door has the prize
})
mean(switch)    # probability of choosing prize door when switching
```

##Assessment Part 1

In the 200m dash finals in the Olympics, 8 runners compete for 3 medals (order matters). In the 2012 Olympics, 3 of the 8 runners were from Jamaica and the other 5 were from different countries. The three medals were all won by Jamaica (Usain Bolt, Yohan Blake, and Warren Weir).

Use the information above to help you answer the following four questions.
```{r}
library(gtools)
library(dplyr)
#How many different ways can the 3 medals be distributed across 8 runners?
medals <- permutations(8,3) %>% nrow()
#How many different ways can the three medals be distributed among the 3 runners from Jamaica?
jamaica <- permutations(3,3) %>% nrow()
#What is the probability that all 3 medals are won by Jamaica?
jamaica/medals
#Run a Monte Carlo simulation on this vector representing the countries of the 8 runners in this race:
runners <- c("Jamaica", "Jamaica", "Jamaica", "USA", "Ecuador", "Netherlands", "France", "South Africa")
B <- 10000
all_jamaica <- replicate(B, {
  results <- sample(runners, 3)
  all(results == "Jamaica")
})
mean(all_jamaica)
```

##Assessment Part 2

A restaurant manager wants to advertise that his lunch special offers enough choices to eat different meals every day of the year. He doesn't think his current special actually allows that number of choices, but wants to change his special if needed to allow at least 365 choices.

A meal at the restaurant includes 1 entree, 2 sides, and 1 drink. He currently offers a choice of 1 entree from a list of 6 options, a choice of 2 different sides from a list of 6 options, and a choice of 1 drink from a list of 2 options.
```{r}
library(gtools)
#How many meal combinations are possible with the current menu?
entree <- combinations(6,1)
sides <- combinations(6,2)
drink <- combinations(2,1)
nrow(entree)*nrow(sides)*nrow(drink)
#How many combinations are possible if he expands his original special to 3 drink options?
drink <- combinations(3,1)
nrow(entree)*nrow(sides)*nrow(drink)
#How many meal combinations are there if customers can choose from 6 entrees, 3 drinks, and select 3 sides from the current 6 options?
sides <- combinations(6,3)
nrow(entree)*nrow(sides)*nrow(drink)
#Write a function that takes a number of entree choices and returns the number of meal combinations possible given that number of entree options, 3 drink choices, and a selection of 2 sides from 6 options.
#Use sapply() to apply the function to entree option counts ranging from 1 to 12.
#What is the minimum number of entree options required in order to generate more than 365 combinations?
menu <- function(entrees) {
  entree <- combinations(entrees,1)
  sides <- combinations(6,2)
  drink <- combinations(3,1)
  nrow(entree)*nrow(sides)*nrow(drink)
}
sapply(seq(1,12),menu)
#- Write a function that takes a number of side choices and returns the number of meal combinations possible given 6 entree choices, 3 drink choices, and a selection of 2 sides from the specified number of side choices.
#Use sapply() to apply the function to side counts ranging from 2 to 12.
menu2 <- function(num_sides) {
  entree <- combinations(6,1)
  sides <- combinations(num_sides,2)
  drink <- combinations(3,1)
  nrow(entree)*nrow(sides)*nrow(drink)
}
sapply(seq(2,12),menu2)
```

##Assessment Part 3

The built-in dataset esoph contains data from a case-control study in France comparing people with esophageal cancer (cases, counted in ncases) to people without esophageal cancer (controls, counted in ncontrols) that are carefully matched on a variety of demographic and medical characteristics. 

Each row contains one group of the experiment. Each group has a different combination of age, alcohol consumption, and tobacco consumption. The number of cancer cases and number of controls (individuals without cancer) are reported for each group.
```{r}
library(tidyverse)
esoph %>% filter(alcgp == "120+") %>% head()
#How many groups are in the study?
nrow(esoph)
#How many cases are there?
all_cases <- sum(esoph$ncases)
#How many controls are there?
all_controls <- sum(esoph$ncontrols)
#What is the probability that a subject in the highest alcohol consumption group is a cancer case?
high_alcgp_cases <- esoph %>% filter(alcgp == "120+") %>% select(ncases) %>% sum()
high_alcgp_controls <- esoph %>% filter(alcgp == "120+") %>% select(ncontrols) %>% sum()
high_alcgp_cases/(high_alcgp_controls+high_alcgp_cases)
##What is the probability that a subject in the lowest alcohol consumption group is a cancer case?
low_alcgp_cases <- esoph %>% filter(alcgp == "0-39g/day") %>% select(ncases) %>% sum()
low_alcgp_controls <- esoph %>% filter(alcgp == "0-39g/day") %>% select(ncontrols) %>% sum()
low_alcgp_cases/(low_alcgp_controls+low_alcgp_cases)
#Given that a person is a case, what is the probability that they smoke 10g or more a day?
all_study <- all_cases + all_controls
smoke_cases <- esoph %>% filter(!tobgp == "0-9g/day") %>% select(ncases) %>% sum()
smoke_cases/all_cases
#Given that a person is a control, what is the probability that they smoke 10g or more a day?
smoke_controls <- esoph %>% filter(!tobgp == "0-9g/day") %>% select(ncontrols) %>% sum()
smoke_controls/all_controls
```

##Assessment Part 4

```{r}
#For cases, what is the probability of being in the highest alcohol group?
p1 <- esoph %>% filter(alcgp == "120+") %>% select(ncases) %>% sum / all_cases
#For cases, what is the probability of being in the highest tobacco group?
p2 <- esoph %>% filter(tobgp == "30+") %>% select(ncases) %>% sum() / all_cases
#For cases, what is the probability of being in the highest alcohol group and the highest tobacco group?
esoph %>% filter(tobgp == "30+" & alcgp == "120+") %>% select(ncases) %>% sum() / all_cases
#For cases, what is the probability of being in the highest alcohol group or the highest tobacco group?
p_both_cases <- esoph %>% filter(tobgp == "30+" | alcgp == "120+") %>% select(ncases) %>% sum() / all_cases
#For controls, what is the probability of being in the highest alcohol group?
p3 <- esoph %>% filter(alcgp == "120+") %>% select(ncontrols) %>% sum / all_controls
#How many times more likely are cases than controls to be in the highest alcohol group?
p1/p3
#For controls, what is the probability of being in the highest tobacco group?
esoph %>% filter(tobgp == "30+") %>% select(ncontrols) %>% sum() / all_controls
#For controls, what is the probability of being in the highest alcohol group and the highest tobacco group?
esoph %>% filter(tobgp == "30+" & alcgp == "120+") %>% select(ncontrols) %>% sum() / all_controls
#For controls, what is the probability of being in the highest alcohol group or the highest tobacco group?
p_both_control <- esoph %>% filter(tobgp == "30+" | alcgp == "120+") %>% select(ncontrols) %>% sum() / all_controls
#How many times more likely are cases than controls to be in the highest alcohol group or the highest tobacco group?
p_both_cases / p_both_control
```
© 2021 GitHub, Inc.
Terms
Privacy
Security
Status
Docs
Contact GitHub
Pricing
API
Training
Blog
About
