---
title: "Section 2 Continuous Probability"
author: "kevinoliva26"
date: "5/14/2021"
output: html_document
---
##Continuous Probability

The CDF is the probability distribution function for continuous variables. For example, to determine the probability that a male student is taller than 70.5 inches given a vector of male heights  x , we can use the CDF:

*Pr(x>70.5)=1−Pr(x≤70.5)=1−F(70.5) *

The probability that an observation is in between two values  a,b  is  F(b)−F(a) .
```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)

F <- function(a) mean(x <= a)
1 - F(70)    # probability of male taller than 70 inches
```

##Theoretical Distribution

<<<<<<< HEAD
=======
*pnorm(a, avg, s)* gives the value of the cumulative distribution function F(a) for the normal distribution defined by average avg and standard deviation s.

We say that a random quantity is normally distributed with average avg and standard deviation s if the approximation pnorm(a, avg, s) holds for all values of a.

If we are willing to use the normal approximation for height, we can estimate the distribution simply from the mean and standard deviation of our values.
If we treat the height data as discrete rather than categorical, we see that the data are not very useful because integer values are more common than expected due to rounding. This is called discretization.
With rounded data, the normal approximation is particularly useful when computing probabilities of intervals of length 1 that include exactly one integer.

Code: Using pnorm() to calculate probabilities
```{r}
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)
head(x)

#We can estimate the probability that a male is taller than 70.5 inches using:

1 - pnorm(70.5, mean(x), sd(x))
```

Code: Discretization and the normal approximation
```{r}
# plot distribution of exact heights in data
plot(prop.table(table(x)), xlab = "a = Height in inches", ylab = "Pr(x = a)")

# probabilities in actual data over length 1 ranges containing an integer
mean(x <= 68.5) - mean(x <= 67.5)
mean(x <= 69.5) - mean(x <= 68.5)
mean(x <= 70.5) - mean(x <= 69.5)

# probabilities in normal approximation match well
pnorm(68.5, mean(x), sd(x)) - pnorm(67.5, mean(x), sd(x))
pnorm(69.5, mean(x), sd(x)) - pnorm(68.5, mean(x), sd(x))
pnorm(70.5, mean(x), sd(x)) - pnorm(69.5, mean(x), sd(x))

# probabilities in actual data over other ranges don't match normal approx as well
mean(x <= 70.9) - mean(x <= 70.1)
pnorm(70.9, mean(x), sd(x)) - pnorm(70.1, mean(x), sd(x))
```
>>>>>>> f0bd94c... Added content


##Probability Density

In R, the probability density function for the normal distribution is given by *dnorm()*. We will see uses of dnorm() in the future.
<<<<<<< HEAD
Note that *dnorm()* gives the density function and *pnorm()* gives the distribution function, which is the integral of the density function.
=======
Note that *dnorm()* gives the density function and *pnorm()* gives the distribution function, which is the integral of the density function.

##Plotting the probability density for the normal distribution

We can use dnorm() to plot the density curve for the normal distribution. dnorm(z) gives the probability density f(z) of a certain z-score, so we can draw a curve by calculating the density over a range of possible values of z.

First, we generate a series of z-scores covering the typical range of the normal distribution. Since we know 99.7% of observations will be within −3≤z≤3, we can use a value of z slightly larger than 3 and this will cover most likely values of the normal distribution. Then, we calculate f(z), which is dnorm() of the series of z-scores. Last, we plot z against f(z).

```{r}
library(tidyverse)
x <- seq(-4, 4, length = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x, f)) +
    geom_line()
```

Note that dnorm() gives densities for the standard normal distribution by default. Probabilities for alternative normal distributions with mean mu and standard deviation sigma can be evaluated with:

*dnorm(z, mu, sigma)*

##Monte Carlo Simulations

rnorm(n, avg, s) generates n random numbers from the normal distribution with average avg and standard deviation s.
By generating random numbers from the normal distribution, we can simulate height data with similar properties to our dataset. Here we generate simulated height data using the normal distribution.

Code: Generating normally distributed random numbers
```{r}
# define x as male heights from dslabs data
library(tidyverse)
library(dslabs)
data(heights)
x <- heights %>% filter(sex=="Male") %>% pull(height)

# generate simulated height data using normal distribution - both datasets should have n observations
n <- length(x)
avg <- mean(x)
s <- sd(x)
simulated_heights <- rnorm(n, avg, s)

# plot distribution of simulated_heights
data.frame(simulated_heights = simulated_heights) %>%
    ggplot(aes(simulated_heights)) +
    geom_histogram(color="black", binwidth = 2)
```

Code: Monte Carlo simulation of tallest person over 7 feet
```{r}
B <- 10000
tallest <- replicate(B, {
    simulated_data <- rnorm(800, avg, s)    # generate 800 normally distributed random heights
    max(simulated_data)    # determine the tallest height
})
mean(tallest >= 7*12)    # proportion of times that tallest person exceeded 7 feet (84 inches)
```

##Other Continuous Distributions

You may encounter other continuous distributions (Student t, chi-squared, exponential, gamma, beta, etc.).
R provides functions for *density (d)*, *quantile (q)*, *probability distribution (p) * and r*andom number generation (r) * for many of these distributions.
Each distribution has a matching abbreviation (for example, norm() or t()) that is paired with the related function abbreviations *(d, p, q, r)* to create appropriate functions.
For example, use rt() to generate random numbers for a Monte Carlo simulation using the Student t distribution.

Code: Plotting the normal distribution with dnorm
```{r}
x <- seq(-4, 4, length.out = 100)
data.frame(x, f = dnorm(x)) %>%
    ggplot(aes(x,f)) +
    geom_line()
```

##Questions 1 and 2: ACT scores, part 1


```{r}
set.seed(16, sample.kind = "Rounding")
act_scores <- rnorm(10000, 20.9,5.7)

#What is the mean of act_scores
mean(act_scores)

#What is the standard deviation of act_scores?
sd(act_scores)

#A perfect score is 36 or greater (the maximum reported score is 36). In act_scores, how many perfect scores are there out of 10,000 simulated tests?
sum(act_scores >=36)

#In act_scores, what is the probability of an ACT score greater than 30?
mean(act_scores > 30)

#In act_scores, what is the probability of an ACT score less than or equal to 10?
mean(act_scores <= 10)

```

##Assessment: Continuous Probability

The ACT is a standardized college admissions test used in the United States. The four multi-part questions in this assessment all involve simulating some ACT test scores and answering probability questions about them.

For the three year period 2016-2018, ACT standardized test scores were approximately normally distributed with a mean of 20.9 and standard deviation of 5.7. (Real ACT scores are integers between 1 and 36, but we will ignore this detail and use continuous values instead.)

First we'll simulate an ACT test score dataset and answer some questions about it.

Set the seed to 16, then use rnorm() to generate a normal distribution of 10000 tests with a mean of 20.9 and standard deviation of 5.7. Save these values as act_scores. You'll be using this dataset throughout these four multi-part questions.

What is the mean of act_scores?
```{r}
set.seed(16, sample.kind = "Rounding")
act_scores <- rnorm(10000,20.9,5.7)
act_average <- mean(act_scores)
act_sd <- sd(act_scores)
```

A perfect score is 36 or greater (the maximum reported score is 36).
In act_scores, how many perfect scores are there out of 10,000 simulated tests?
```{r}
sum(act_scores>=36)
```

In act_scores, what is the probability of an ACT score greater than 30?
```{r}
1-pnorm(30,act_average,act_sd)
```

In act_scores, what is the probability of an ACT score less than or equal to 10?
```{r}
pnorm(10,act_average,act_sd)
```

Set x equal to the sequence of integers 1 to 36. Use dnorm to determine the value of the probability density function over x given a mean of 20.9 and standard deviation of 5.7; save the result as f_x. Plot x against f_x.
```{r}
x <- seq(1,36)
f_x <- dnorm(x,20.9,5.7)
plot(x,f_x)
```

Convert act_scores to Z-scores. Recall from Data Visualization (the second course in this series) that to standardize values (convert values into Z-scores, that is, values distributed with a mean of 0 and standard deviation of 1), you must subtract the mean and then divide by the standard deviation. Use the mean and standard deviation of act_scores, not the original values used to generate random test scores.

What is the probability of a Z-score greater than 2 (2 standard deviations above the mean)?
```{r}
act_z_score <- (act_scores - act_average)/act_sd
act_z_average <- mean(act_z_score)
act_z_sd <- sd(act_z_score)
1-pnorm(2,act_z_average,act_z_sd)
```

What ACT score value corresponds to 2 standard deviations above the mean (Z = 2)?
```{r}
qnorm(pnorm((act_average + 2 *act_sd),act_average,act_sd),act_average,act_sd)

#or easier with this
2*sd(act_scores) + mean(act_scores)
```

A Z-score of 2 corresponds roughly to the 97.5th percentile.
Use qnorm() to determine the 97.5th percentile of normally distributed data with the mean and standard deviation observed in act_scores.
What is the 97.5th percentile of act_scores?
```{r}
qnorm(0.975, act_average,act_sd)
```

Write a function that takes a value and produces the probability of an ACT score less than or equal to that value (the CDF). Apply this function to the range 1 to 36.

```{r}
#Option 1
prob_act <- function(a){
    pnorm(a,act_average,act_sd)
}
sapply(seq(1:36),prob_act)
qnorm(sapply(seq(1:36),prob_act),act_average,act_sd)

qnorm(0.9632900984,act_average,act_sd)

#easier option
cdf <- sapply(1:36, function (x){
  mean(act_scores <= x)
})
min(which(cdf >= .95))
```

Use qnorm() to determine the expected 95th percentile, the value for which the probability of receiving that score or lower is 0.95, given a mean score of 20.9 and standard deviation of 5.7.

What is the expected 95th percentile of ACT scores?
```{r}
qnorm(0.95,20.9,5.7)
```

As discussed in the Data Visualization course, we can use quantile() to determine sample quantiles from the data.
Make a vector containing the quantiles for p <- seq(0.01, 0.99, 0.01), the 1st through 99th percentiles of the act_scores data. Save these as sample_quantiles.
In what percentile is a score of 26?
```{r}
p <- seq(0.01, 0.99, 0.01)
sample_quantiles <- quantile(act_scores, p)
names(sample_quantiles[max(which(sample_quantiles < 26))])
```

Make a corresponding set of theoretical quantiles using qnorm() over the interval p <- seq(0.01, 0.99, 0.01) with mean 20.9 and standard deviation 5.7. Save these as theoretical_quantiles. Make a QQ-plot graphing sample_quantiles on the y-axis versus theoretical_quantiles on the x-axis.
```{r}
library(ggplot2)
theorical_quantiles <- qnorm(p,20.9,5.7)
qplot(theorical_quantiles,sample_quantiles) + geom_abline()
```
