---
title: "Section 4 The Big Short"
author: "kevinoliva26"
date: "6/27/2021"
output: html_document
---

##The Big Short: Interest Rates Explained

Interest rates for loans are set using the probability of loan defaults to calculate a rate that minimizes the probability of losing money.
We can define the outcome of loans as a random variable. We can also define the sum of outcomes of many loans as a random variable.
The Central Limit Theorem can be applied to fit a normal distribution to the sum of profits over many loans. We can use properties of the normal distribution to calculate the interest rate needed to ensure a certain probability of losing money for a given probability of default.

Code: Interest rate sampling model
```{r}
n <- 1000
loss_per_foreclosure <- -200000
p <- 0.02
defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE)
sum(defaults * loss_per_foreclosure)
```

Code: Interest rate Monte Carlo simulation
```{r}
B <- 10000
losses <- replicate(B, {
    defaults <- sample( c(0,1), n, prob=c(1-p, p), replace = TRUE) 
  sum(defaults * loss_per_foreclosure)
})
```

Code: Plotting expected losses
```{r}
library(tidyverse)
data.frame(losses_in_millions = losses/10^6) %>%
    ggplot(aes(losses_in_millions)) +
    geom_histogram(binwidth = 0.6, col = "black)"
```

Code: Expected value and standard error of the sum of 1,000 loans
```{r}
n*(p*loss_per_foreclosure + (1-p)*0)    # expected value 
sqrt(n)*abs(loss_per_foreclosure)*sqrt(p*(1-p))    # standard error
```

Code: Calculating interest rates for expected value of 0
We can calculate the amount  to add to each loan so that the expected value is 0 using the equation *lp + x(1-p) = 0*. Note that this equation is the definition of expected value given a loss per foreclosure  with foreclosure probability  and profit  if there is no foreclosure (probability 1-p).

We solve for x = -(lp)/(1-p) and calculate :
```{r}
x = - loss_per_foreclosure*p/(1-p)
x
#On a $180,000 loan, this equals an interest rate of:
x/180000
```


Code: Calculating interest rate for 1% probability of losing money
```{r}
l <- loss_per_foreclosure
z <- qnorm(0.01)
x <- -l*( n*p - z*sqrt(n*p*(1-p)))/ ( n*(1-p) + z*sqrt(n*p*(1-p)))/x
x/180000    # interest rate
loss_per_foreclosure*p + x*(1-p)    # expected value of the profit per loan
n*(loss_per_foreclosure*p + x*(1-p)) # expected value of the profit over n loans
```

Code: Monte Carlo simulation for 1% probability of losing money
```{r}
B <- 100000
profit <- replicate(B, {
    draws <- sample( c(x, loss_per_foreclosure), n, 
                        prob=c(1-p, p), replace = TRUE) 
    sum(draws)
})
mean(profit)    # expected value of the profit over n loans
mean(profit<0)    # probability of losing money
```











